---
title: "Time Series Modeling - Bihag"
author: "Raphael Bihag"
date: "`r Sys.Date()`"
output: 
  pdf_document: 
    latex_engine: xelatex
---

# Time Series Modeling

## Introduction

Enter Short Description hereeee

## Data Wrangling

### Load Required Libraries
```{r load_packages, message=FALSE, warning=FALSE}
library(forecast)
library(tseries)
library(ggplot2)
library(urca)
library(TSA)
library(dplyr)
```

### Assess the Data
The data we will be using for this session is Lake Huron data that is available natively in R. Let's plot it and get its summary.

```{r load_dataset}
# Loads dataset to a variable
lakehuron <- LakeHuron

# plots the dataset
plot(lakehuron, main="Lake Huron Time Series", xlab="Year", ylab="Level (ft)")

# Prints out a few lines of the dataset
lakehuron
```
The output shows that the data is yearly (frequency=1) and ranges from 1875 - 1972. This data is already cleaned and ready for analysis. Before doing our time series modeling, we have to test for the data's stationarity.

### Stationarity Test & Transformation
We are going to test for the data's stationarity using the Augmented Dickey-Fuller test, which tests the null hypothesis that the data is non-stationary.

```{r stationarity_test}
adf.test(lakehuron)
```
The result of the Augmented Dickey-Fuller test shows that the p-value is 0.254 which is greater than the 5% level of significance. Thus, we fail to reject the null hypothesis; the dataset is non-stationary.

Because the dataset is non-stationary, we perform transformations to it for it to be stationary. For this case, we difference the dataset and check if it becomes stationary.

```{r transformation}
lakehuron_diff <- diff(lakehuron, lag=1)
adf.test(lakehuron_diff)
```
The data has become stationary after first iteration of differencing. Thus, the correct level of integration for this data is 1. We are now set for analysis.

## Model Identification
In this section, we perform ACF/PACF to identify the significant lag order for AR and MA components 
of the models. This step is important because it tells us what $p$ and $q$ are for when modeling $AR(p)$ and $MA(q)$.

```{r ACF_plotting}
acf(lakehuron_diff, lag.max = 30, main="ACF 1st Differenced Lake Huron")
pacf(lakehuron_diff, lag.max = 30, main="PACF 1st Differenced Lake Huron")
```

```{r identify_lag}
# computes the confidence bound for the dataset
conf_bound <- 1.96 / sqrt(length(lakehuron_diff))

# stores the acf plot data
acf_vals <- acf(lakehuron_diff, plot = FALSE, lag.max = 30)
pacf_vals <- pacf(lakehuron_diff, plot = FALSE, lag.max = 30)

# identi
sig_acf <- which(abs(acf_vals$acf[-1]) > conf_bound)
sig_pacf <- which(abs(pacf_vals$acf) > conf_bound)
```

