---
title: "Time Series Modeling - Bihag"
author: "Raphael Bihag"
date: "`r Sys.Date()`"
output: 
  pdf_document: 
    latex_engine: xelatex
---

# Time Series Modeling

## Introduction

Enter Short Description hereeee

## Data Wrangling

### Load Required Libraries
```{r load_packages, message=FALSE, warning=FALSE}
library(forecast)
library(tseries)
library(ggplot2)
library(urca)
library(TSA)
library(dplyr)
```

### Assess the Data
The data we will be using for this session is Lake Huron data that is available natively in R. Let's plot it and get its summary.

```{r load_dataset}
# Loads dataset to a variable
lakehuron <- LakeHuron

# plots the dataset
plot(lakehuron, main="Lake Huron Time Series", xlab="Year", ylab="Level (ft)")

# Prints out a few lines of the dataset
lakehuron
```
The output shows that the data is yearly (frequency=1) and ranges from 1875 - 1972. This data is already cleaned and ready for analysis. Before doing our time series modeling, we have to test for the data's stationarity.

### Stationarity Test & Transformation
We are going to test for the data's stationarity using the Augmented Dickey-Fuller test, which tests the null hypothesis that the data is non-stationary.

```{r stationarity_test}
adf.test(lakehuron)
```
The result of the Augmented Dickey-Fuller test shows that the p-value is 0.254 which is greater than the 5% level of significance. Thus, we fail to reject the null hypothesis; the dataset is non-stationary.

Because the dataset is non-stationary, we perform transformations to it for it to be stationary. For this case, we difference the dataset and check if it becomes stationary.

```{r transformation}
lakehuron_diff <- diff(lakehuron, lag=1)
adf.test(lakehuron_diff)
```
The data has become stationary after first iteration of differencing. Thus, the correct level of integration for this data is 1. We are now set for analysis.

## Model Identification
In this section, we perform ACF/PACF to identify the significant lag order for AR and MA components 
of the models. This step is important because it tells us what $p$ and $q$ are for when modeling $AR(p)$ and $MA(q)$.

```{r ACF_plotting}
acf(lakehuron_diff, main="ACF Lake Huron (1st Differenced)", lag.max = 30)
pacf(lakehuron_diff, main="PACF Lake Huron (1st Differenced)", lag.max = 30)
```

```{r identify_lag}
# computes the confidence bound for the dataset
conf_bound <- 1.96 / sqrt(length(lakehuron_diff))

# stores the acf plot data
acf_vals <- acf(lakehuron_diff, plot = FALSE, lag.max = 30)
pacf_vals <- pacf(lakehuron_diff, plot = FALSE, lag.max = 30)

# identifies which lags are significant from ACF and PACF
sig_acf <- which(abs(acf_vals$acf[-1]) > conf_bound)
sig_pacf <- which(abs(pacf_vals$acf) > conf_bound)

cat("Significant lags ACF: ", sig_acf, "\n")
cat("Significant lags PACF: ", sig_pacf, "\n")
```
The results show that the significant lags from ACF ($q$) are 2 and 8, and that from PACF ($p$) are 2 and 20. With these values, we define our significant lag order for AR and MA components as variables to be used for later. We will also define the order of integration ($d$) as 1 since we differenced the data once to make it stationary.

```{r lag_orders}
ar_order <- 20
ma_order <- 8
d_value <- 1
```


## Model Fitting
Before fitting our models, we have to split the data into train set and test set at 80-20 ratio.

### Data Splitting

```{r data_split}
n <- length(lakehuron)
train_size <- floor(0.8 * n)
lh_train <- window(lakehuron, end = time(lakehuron)[train_size])
lh_test <- window(lakehuron, start = time(lakehuron)[train_size + 1])

cat("Lake Huron Split \n")
cat("> Training Set:", length(lh_train), "obs (", start(lh_train)[1], "-", end(lh_train)[1], ")\n")
cat("> Test Set:", length(lh_test), "obs (", start(lh_test)[1], "-", end(lh_test)[1], ")\n")
```
### Model Fitting
Next, we autofit an ARIMA model using the AR and MA orders identified. The model is defined as $ARIMA(p, d, q)$ where $p$ is the number of significant lags of PACF, $q$ is the number of significant lags of ACF, and $d$ is the order of integration.

```{r model_fitting}
# Variable to hold all of the results
results <- list()

# fitting the AR model
results$ar <- Arima(lh_train, order = c(ar_order, d_value, 0))

# fitting the MA model
results$ma <- Arima(lh_train, order = c(0, d_value, ma_order))

# fitting the ARMA model
results$arma <- Arima(lh_train, order = c(ar_order, d_value, ma_order))

# fitting the auto-ARIMA model
results$auto <- auto.arima(lh_train)
```


### Forecasting
We have successfully fitted all our models. Next, let's forecast using these models, which we will use later to evaluate each model's accuracy. The forecast duration must be the same length as the test data.

```{r forecasting}
forecasts <- list()

# Forecasting using AR
forecasts$ar <- forecast(results$ar, h = length(lh_test))

# Forecasting using MA
forecasts$ma <- forecast(results$ma, h = length(lh_test))

# Forecasting using ARMA
forecasts$arma <- forecast(results$arma, h = length(lh_test))

# Forecasting using auto-ARIMA
forecasts$auto <- forecast(results$auto, h = length(lh_test))
```

```{r forecast_plot}
par(mfrow = c(2,2))
plot(forecasts$ar, main= "AR", ylim=range(lakehuron))
lines(lh_test,col= "red")

plot(forecasts$ma, main= "MA", ylim=range(lakehuron))
lines(lh_test,col= "red")

plot(forecasts$arma, main= "ARMA", ylim=range(lakehuron))
lines(lh_test,col= "red")

plot(forecasts$auto, main= "Auto-ARIMA", ylim=range(lakehuron))
lines(lh_test,col= "red")
```



