---
title: "Time Series Foundations - Bihag"
author: "Raphael Bihag"
date: "`r Sys.Date()`"
output: 
  pdf_document: 
    latex_engine: xelatex
---

# Introduction

# Part 0: Setting up RMarkdown in RStudio

## Step 1: Create RMarkdown File
1. Click File → New File → R Markdown
2. Title: Time Series Foundations
3. Output: HTML (recommended)
4. Click OK


## Step 2: Installing the required packages
Note: you only need to do this once.

install.packages("forecast")
install.packages("tseries")
install.packages("ggplot2")
install.packages("urca")
install.packages("TSA")


## Step 3: Understanding RMarkdown Chunks.
RMarkdowns allows combining of text+code+output

```{r chunk_name}

# Your R code here. (The hashtag means it is a comment)

```

Keyboard Shortcut:

Mac: Command + Option + I

Windows: Ctrl + Alt + I

# Part 1: Loading Libraries
Create a chunk named setup and load the libraries to be used.

```{r setup}
library(forecast)
library(tseries)
library(ggplot2)
library(urca)
library(TSA)
```

# Part 2: Overview of Learning Goals
In this practical session, the students will use available dataset in R to:

1. Create Time Series Graphics
2. Perform Statistical Tests for Trend, Drift, Seasonality
3. Perform ADF Stationarity Test
4. Interpret results

# Example 1: Air Passengers Dataset
Data Info: Monthly Airline Passenger Numbers 1949-1960. The classic Box & Jenkins airline data. Monthly totals of international airline passengers, 1949 to 1960.

```{r example}
# Step 1: Load dataset
data("AirPassengers")
ap <- AirPassengers

# Step 2: Time series plot
plot(ap, main="AirPassengers Time Series")

# Step 3: Decomposition (Trend + seasonality Test)
decomp_ap <- decompose(ap)
plot(decomp_ap)

# Step 4: Drift Test (Mean Change Over Time)
t.test(window(ap, end = c(1955,12)),
window(ap, start = c(1956,1)))

# Step 5: ADF Test
adf.test(ap)
```
```{r example_transformation}
# Step 6: Data Transformations

# Step 6.1: Log transform the data and test
log_ap <- log(ap)
adf.test(log_ap)

# Step 6.2: Difference the data and test
d1_ap <- diff(ap)
adf.test(d1_ap)
```
# Example 2: LakeHuron Dataset
Data Info: Level of Lake Huron 1875–1972.
Annual measurements of the level, in feet, of Lake Huron 1875–1972.

```{r example2_LakeHuron}
# Step 1: Load dataset and plot the series
data("LakeHuron")
lh <- LakeHuron
plot(lh, main="Lake Huron Water Levels")

# Step 2. ADF test.
adf.test(lh)

# Step 3: Data Transformation (Log)
log_lh <- log(lh)
plot(log_lh, main="Log-transformed Lake Huron Water Levels")
adf.test(log_lh)

# Step 4: Data transformation (difference)
dl_loglh <- diff(log_lh)
plot(dl_loglh, main="1st Difference of Log-transformed Lake Huron Water Levels")
adf.test(dl_loglh)
```

# Practical Session 1

Data Info: Daily Closing Prices of Major European Stock Indices, 1991–1998
Contains the daily closing prices of major European stock indices: Germany DAX (Ibis),
Switzerland SMI, France CAC, and UK FTSE. The data are sampled in business time, i.e.,
weekends and holidays are omitted.

## Loading dataset

```{r load_eu}
data("EuStockMarkets")
eu <- EuStockMarkets[, "DAX"]
plot(eu, main="EU Stock Markets (DAX)")
```

## Stationarity Testing
We can test if the dataset is stationary or not using the Augmented Dickey-Fuller test. This tests the data against the null hypothesis of it being nonstationary. 

```{r eu_stationarity}
adf.test(eu)
```
With the p-value being more than than 0.05, we fail to reject the null hypothesis---the data is found to be stationary.

## Time Series Decomposition
```{r decomposition_eu}
decomp_eu <- decompose(eu)
plot(decomp_eu)
```

From the Time Series Decomposition graph, there seems to be an upward trend going on, with a steepness increasing from around 1997 to 1998. Seasonality can also be observed, wherein it dips around the end of the year and peaks at around 2/3.

The ADF test performed found the data to be non-stationary. Because of this, data transformations must be made to continue our analysis.

## Data Transformation

Transformations to the data can help it be stationary. The two common transformation methods are Log Transformation and Differencing. The former is great in helping stabilizing variance. This method is for data whose variance appear to be changing over time (data becomes "wider"). The latter is great for removing the trend and seasonality of the data. This is great for data that exhibit both.

For our data (EU Stock Market-DAX), there appears to be no changes in variance over time, only trend and seasonality. Because of this, differencing is deemed to be the appropriate method to making the data stationary.

```{r eu_diff}
eu_diff <- diff(eu)
plot(eu_diff, main="1st Difference of EU Stock Market-DAX")
adf.test(eu_diff)
```
After the transformation, the ADF test resulted a p-value<0.05, thus we reject the null hypothesis. The 1st differenced data is stationary.